### 2025-07-22 paper/quarto_aaai/paper.tex

\caption{\label{tbl-main}Key evaluation metrics for valid counterfactual
along with bootstrapped standard errors for all datasets.
\textbf{Plausibility} (columns 1-2): percentage reduction in
implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
\textbf{Cost} / \textbf{Actionability} (column 3): percentage reduction
in costs when selected features are protected. Outcomes are aggregated
across bootstrap samples and varying degrees of the energy penalty
\(\lambda_{\text{egy}}\) used for \emph{ECCCo} at test time. Asterisks
(\(^*\)) indicate that the bootstrapped 99\%-confidence interval of
differences in mean outcomes does \emph{not} include zero.}

>>>>

\caption{\label{tbl-main}Key evaluation metrics for valid counterfactual
along with bootstrapped standard errors for all datasets.
\textbf{Plausibility} (columns 1-2): percentage reduction in
implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
\textbf{Cost} / \textbf{Actionability} (column 3): percentage reduction
in costs when selected features are protected. Outcomes are aggregated
across bootstrap samples (100 rounds) and varying degrees of the energy
penalty \(\lambda_{\text{egy}}\) used for \emph{ECCCo} at test time.
Asterisks (\(^*\)) indicate that the bootstrapped 99\%-confidence
interval of differences in mean outcomes does \emph{not} include zero.}

### 2025-07-22 paper/sections/abstract.qmd

We propose a novel training regime termed counterfactual training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the feature mutability constraints. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations and have improved adversarial robustness.

>>>>

We propose a novel training regime termed counterfactual training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the feature mutability constraints. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable counterfactual explanations and exhibit greatly improved adversarial robustness.

### 2025-07-22 paper/sections/discussion.qmd

*Plausibility is costly.* As noted elsewhere [@altmeyer2024faithful], more plausible counterfactuals are inevitably more costly. Thus, while CT improves plausibility and robustness, it can impact average costs and validity as cheap, implausible and adversarial explanations are removed from the solution space.

*CT increases the training times.* Like contrastive and robust learning, CT is more resource-intensive than conventional regimes. Three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.

>>>>

*Plausibility is costly.* As noted by @altmeyer2024faithful, more plausible counterfactuals are inevitably more costly. CT improves plausibility and robustness, but it can impact average costs and validity when cheap, implausible and adversarial explanations are removed from the solution space.

*CT increases the training times.* Just like contrastive and robust learning, CT is more resource-intensive than conventional regimes. Three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.

### 2025-07-22 paper/sections/experiments.qmd

::: {#tbl-main}
```{=latex}
\begin{table}[h]
\small
\centering
\begin{tabular}{
  l
  S[table-format=2.2(1.2)]
  S[table-format=3.2(3.2)]
  S[table-format=3.2(1.2)]
}
  \toprule
  \textbf{Data} & \textbf{$ \text{IP} $ $(-\%)$} & \textbf{$ \text{IP}^* $ $(-\%)$} & \textbf{Cost $(-\%)$} \\\midrule
  LS & 29.05\pm0.67 $^{*}$ & 55.33\pm2.03 $^{*}$ & 14.07\pm0.6 $^{*}$ \\
  Circ & 56.29\pm0.44 $^{*}$ & 89.38\pm9.3 $^{*}$ & 45.55\pm0.76 $^{*}$ \\
  Moon & 20.62\pm0.69 $^{*}$ & 19.26\pm8.12 $^{*}$ & 2.86\pm1.03 $^{*}$ \\
  OL & -1.13\pm0.88 $^{}$ & -24.52\pm14.52 $^{}$ & 38.39\pm2.21 $^{*}$ \\\midrule
  Adult & 0.77\pm1.34 $^{}$ & 32.29\pm6.87 $^{*}$ & -2.82\pm4.88 $^{}$ \\
  CH & 12.05\pm1.41 $^{*}$ & 70.27\pm3.72 $^{*}$ & 40.71\pm1.55 $^{*}$ \\
  Cred & 12.31\pm1.84 $^{*}$ & 54.89\pm11.21 $^{*}$ & -17.43\pm5.17 $^{*}$ \\
  GMSC & 23.44\pm1.99 $^{*}$ & 73.31\pm4.83 $^{*}$ & 62.64\pm2.04 $^{*}$ \\
  MNIST & 7.05\pm1.8 $^{*}$ & -25.09\pm109.05 $^{}$ & -12.34\pm6.52 $^{}$ \\\midrule
  Avg. & 17.83 & 38.35 & 19.07 \\\bottomrule
\end{tabular}
\end{table}
```
Key evaluation metrics for valid counterfactual along with bootstrapped standard errors for all datasets. **Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Cost** / **Actionability** (column 3): percentage reduction in costs when selected features are protected. Outcomes are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCCo* at test time. Asterisks ($^*$) indicate that the bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.
:::

>>>>

::: {#tbl-main}
```{=latex}
\begin{table}[h]
\small
\centering
\begin{tabular}{
  l
  S[table-format=2.2(1.2)]
  S[table-format=3.2(3.2)]
  S[table-format=3.2(1.2)]
}
  \toprule
  \textbf{Data} & \textbf{$ \text{IP} $ $(-\%)$} & \textbf{$ \text{IP}^* $ $(-\%)$} & \textbf{Cost $(-\%)$} \\\midrule
  LS & 29.05\pm0.67 $^{*}$ & 55.33\pm2.03 $^{*}$ & 14.07\pm0.6 $^{*}$ \\
  Circ & 56.29\pm0.44 $^{*}$ & 89.38\pm9.3 $^{*}$ & 45.55\pm0.76 $^{*}$ \\
  Moon & 20.62\pm0.69 $^{*}$ & 19.26\pm8.12 $^{*}$ & 2.86\pm1.03 $^{*}$ \\
  OL & -1.13\pm0.88 $^{}$ & -24.52\pm14.52 $^{}$ & 38.39\pm2.21 $^{*}$ \\\midrule
  Adult & 0.77\pm1.34 $^{}$ & 32.29\pm6.87 $^{*}$ & -2.82\pm4.88 $^{}$ \\
  CH & 12.05\pm1.41 $^{*}$ & 70.27\pm3.72 $^{*}$ & 40.71\pm1.55 $^{*}$ \\
  Cred & 12.31\pm1.84 $^{*}$ & 54.89\pm11.21 $^{*}$ & -17.43\pm5.17 $^{*}$ \\
  GMSC & 23.44\pm1.99 $^{*}$ & 73.31\pm4.83 $^{*}$ & 62.64\pm2.04 $^{*}$ \\
  MNIST & 7.05\pm1.8 $^{*}$ & -25.09\pm109.05 $^{}$ & -12.34\pm6.52 $^{}$ \\\midrule
  Avg. & 17.83 & 38.35 & 19.07 \\\bottomrule
\end{tabular}
\end{table}
```

Key evaluation metrics for valid counterfactual along with bootstrapped standard errors for all datasets. **Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Cost** / **Actionability** (column 3): percentage reduction in costs when selected features are protected. Outcomes are aggregated across bootstrap samples (100 rounds) and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCCo* at test time. Asterisks ($^*$) indicate that the bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.
:::

### 2025-07-22 paper/sections/related.qmd

In a closely related work, @altmeyer2024faithful show that model averaging and, in particular, contrastive model objectives can produce more explainable and hence trustworthy models. The authors propose a way to generate counterfactuals that are maximally faithful models in that they are consistent with what models have learned about the underlying data. Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the contrastive divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by a model. Their counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them. The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your], a form of constrastive learning, tend to do well in this regard.

>>>>

In a closely related work, @altmeyer2024faithful show that model averaging and, in particular, contrastive model objectives can produce more explainable and hence trustworthy models. The authors propose a way to generate counterfactuals that are maximally faithful in that they are consistent with what models have learned about the underlying data. Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the contrastive divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by a model. Their algorithm, *ECCCo*, yields plausible counterfactual explanations if and only if the underlying model has learned representations that align with them. The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your], a form of constrastive learning, tend to do well in this regard.
